---
title: "cosmx_sccoda"
output: html_document
date: "2024-06-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Cell type proportions for spatial data (CosMx)

# Load libraries 
```{r}
library(dplyr)
library(Seurat)
library(SeuratObject)
library(hdf5r)
library(ggplot2)
library(tidyr)
library(fgsea)
library(SeuratData)
library(SeuratDisk)
```

## Load cosmx seurat object 
```{r}
cosmx <- readRDS(here::here("Data", "cosmx.rds"))
```


## Add meta data with correct column names 
```{r}
# load fov annotations 
fov_annotations <- read.csv(here::here("Data", "fov_annotations.csv"))
fov_annotations$fov <- as.character(fov_annotations$fov)

# Extract the fov information from the Seurat object
fov_info <- sub(".*_(.*)", "\\1", colnames(cosmx))

# Add the fov info as a new column in the Seurat object's metadata
cosmx <- AddMetaData(cosmx, fov_info, col.name = "fov")

# Join the fov annotations to get sample IDs
# Convert to data frame for easy manipulation
metadata_df <- cosmx@meta.data %>%
  left_join(fov_annotations, by = c("fov" = "fov"))

# Ensure that the new metadata column is added to the Seurat object
cosmx@meta.data <- metadata_df

names(cosmx@meta.data)[names(cosmx@meta.data) == "id"] <- "SampleID"
names(cosmx@meta.data)[names(cosmx@meta.data) == "sex"] <- "sex"
names(cosmx@meta.data)[names(cosmx@meta.data) == "drug"] <- "condition"
```

## Save seurat object as h5ad for sccoda analysis in python 
```{r}
# v5 of seurat makes the RNA assay at assayv5 but if you need to convert to anndata for sccoda it needs to be "assay" and not assayv5 hence run: 
cosmx@assays$RNA <- as(object = cosmx@assays$RNA, Class = "Assay")
cosmx@assays

# Save object as h5ad for input into sccoda 
SaveH5Seurat(cosmx, filename = here::here("Data", "cosmx_metadata.h5Seurat"), overwrite = TRUE)
Convert(here::here("Data", "cosmx_metadata.h5Seurat"), dest = "h5ad", overwrite = TRUE)

# Save gex_atac as rds 
saveRDS(cosmx, file = here::here("Data", "cosmx_metadata.rds"))
```


## h5ad inputted into python for sccoda 

## PYTHON CODE 

# scCODA: Compositional Analysis for CosMx Spatial Data 

## Import libs
```{python}
import warnings
warnings.filterwarnings("ignore")

import mudata as mu
import matplotlib.pyplot as plt
import pertpy as pt
```


## Import libs 
```{python setup}
# Imports
import pandas as pd
import matplotlib.pyplot as plt
import anndata as ad
import warnings
#from pertpy.sccoda.util import cell_composition_data as dat
#from sccoda.util import data_visualization as viz

#import sccoda.datasets as scd

warnings.filterwarnings("ignore")
```

## Import data 
```{python read_ad}
adata = ad.read_h5ad("cosmx_metadata.h5ad")
print(adata)

```

## Change saline to "Control" as per requirements of scCODA 
```{python}
def control_fun(x):
    if x == "Saline":
      return "Control"
    else:
     return "Fentanyl"

adata.obs['condition'] = adata.obs['condition'].apply(control_fun)

```


# run SCCODA
```{python}
sccoda_model = pt.tl.Sccoda()
sccoda_data = sccoda_model.load(adata, type="cell_level", generate_sample_level=True, cell_type_identifier="aba_classification", sample_identifier="SampleID", covariate_obs=["condition", "sex", "aba_classification"])
print(sccoda_data)
print(sccoda_data["coda"].X)
print(sccoda_data["coda"].obs)
```

## How fentanyl affects cell composition 
```{python}
# Select saline and fentanyl data
sccoda_data.mod["coda_fent"] = sccoda_data["coda"][sccoda_data["coda"].obs["condition"].isin(["Control", "Fentanyl"])].copy()
print(sccoda_data["coda_fent"])
```

```{python}
pt.pl.coda.boxplots(sccoda_data, modality_key="coda_fent", feature_name="condition", add_dots=True)
plt.show()
plt.savefig('boxplot_cosmx.png', bbox_inches='tight')
```

```{python}
# Model set up and inference 

sccoda_data = sccoda_model.prepare(sccoda_data, modality_key="coda_fent", formula="condition", reference_cell_type="automatic")
sccoda_data["coda_fent"]
```

```{python}
# Run MCMC
sccoda_model.run_nuts(sccoda_data, modality_key="coda_fent")
sccoda_data["coda_fent"]
```

```{python}
sccoda_model.set_fdr(sccoda_data, modality_key="coda_fent", est_fdr=0.4)
```

## Save condition[T.fentanyl] as a csv 
```{python}
df = sccoda_data["coda_fent"].varm['effect_df_condition[T.Fentanyl]'] 
df.to_csv('fent_cosmx.csv', index=False)
```


```{python}
#Result Interpretation 
sccoda_model.summary(sccoda_data, modality_key="coda_fent")
# Save summary?
sim_results = sccoda_model.summary(sccoda_data, modality_key="coda_fent")
```

```{python}
sccoda_model.credible_effects(sccoda_data, modality_key="coda_fent")
```

```{python}
# Stacked barplot for the levels of "Condition"
pt.pl.coda.stacked_barplot(sccoda_data, modality_key="coda_fent", feature_name="condition")
plt.show()
plt.savefig('stacked_barplot_cosmx.png', bbox_inches='tight')
```


