---
title: "seurat_clustering.Rmd"
output: html_document
date: "2024-05-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Seurat Clustering of GEX data and MultiVI (GEX + ATAC) pre-processing 

## Load Libraries
```{r}
library(dplyr)
library(Seurat)
library(SeuratObject)
library(hdf5r)
library(ggplot2)
library(tidyr)
library(fgsea)
library(SeuratData)
library(SeuratDisk)
```


## Load Data 
```{r}
# Load rds from QC metrics step 
RNA <- readRDS(here::here("Data", "RNA_subset.rds"))
```

## Normalizing the data 
```{r}
#Normalizing the Data 
RNA <- NormalizeData(RNA, normalization.method = "LogNormalize", scale.factor = 
10000)
```

## Identification of highly variable features 
```{r}
RNA <- FindVariableFeatures(RNA, selection.method = "vst", nfeatures = 2000)
top10 <- head(VariableFeatures(RNA), 10)
plot1 <- VariableFeaturePlot(RNA)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot1 + plot2

```

## Variance stablization and transformation 
```{r}
RNA <- SCTransform(RNA)
```

## Performing linear dimensional reduction 
```{r}
#Performing PCA on the scaled dataset 
RNA <- RunPCA(RNA)

VizDimLoadings(RNA, dims = 1:2, reduction = "pca")
DimPlot(RNA, reduction = "pca")

# Determine dimensionality of the dataset 
ElbowPlot(RNA)
```

## Cluster the cells 
```{r}
RNA <- FindNeighbors(RNA, dims = 1:11)
RNA <- FindClusters(RNA, resolution = 0.1)

```

## Run non-linear dimensional reduction (UMAP)
```{r}
#(reticulate::py_install(packages = 'umap-learn')
set.seed(10)
RNA <- RunUMAP(RNA, dims = 1:11)
DimPlot(RNA, reduction = "umap", label = FALSE)
#Cluster by sample 
DimPlot(RNA, reduction = "umap", group.by = "orig.ident")
```

## Manual annotation 
```{r}

#Now let's extract the top marker genes, and see which ones correspond with each cluster. This can be done using the FindAllMarkers function within Seurat.

RNA <- PrepSCTFindMarkers(object = RNA, assay = "SCT")
cluster_markers <- FindAllMarkers(RNA) # finds which genes are different 
```

## Fast gene set enrichment analysis (fgsea) for only GEX data 
```{r}
mouse_brain <- read.csv(here::here("Data", "biorxiv_mouse_brain.csv"))
mbrain_genesets <- sapply(strsplit(mouse_brain$Optimally.Sized.Gene.List, "\\||="), function(i){
  i[-grep("ENS", i)]
})
names(mbrain_genesets) <- mouse_brain$Cell.Type.Name
mbrain_genesets <- mbrain_genesets[sapply(mbrain_genesets, length) != 0]

fc <- cluster_markers$avg_log2FC
names(fc) <- rownames(cluster_markers)
fc_cl <- split(fc, cluster_markers$cluster)

cluster_annotations <- lapply(fc_cl, function(i){
  res <- fgsea(mbrain_genesets, i)
  res <- res[order(res$pval),]
  res$pathway[1]
})
```

## Labelling clusters with cell type 
```{r}
RNA <- RenameIdents(RNA, cluster_annotations)
DimPlot(RNA, reduction = 'umap')
```

## Save GEX seurat object
```{r}
saveRDS(RNA, file = here::here("Data", 'GEX_seur.rds'))
```


## Add data (clusters) from MultiVI (GEX + ATAC) into seurat object 
```{r}
#read in csv 
leiden_data <- read.csv(here::here("Data", "leiden_column_final.csv"))

#convert to factor class 
leiden_data_factor <- factor(leiden_data$leiden)

# replace "seurat_clusters" in metadata with leiden_data 
RNA@meta.data$seurat_clusters <- leiden_data_factor

# replace Idents in metadata with leiden_data 
leiden_column <- leiden_data$leiden
Idents(RNA) <- leiden_column
```

## Find Cluster Markers 
```{r}
# find markers for each cluster 
cluster_markers <- FindAllMarkers(RNA)

# order the clusters
cluster_markers$cluster <- factor(cluster_markers$cluster, levels = 0:16)

# Arrange the data frame by the ordered 'cluster' column
cluster_markers <- cluster_markers %>%
  arrange(cluster)

# arrange by adj p value and take top 30 genes for each cluster 
cluster_markers_filtered <- cluster_markers %>%  group_by(cluster) %>%  arrange(p_val_adj) %>%  slice_head(n = 150) %>%  ungroup()

# create UpSet plot to make sure no genes are overlapping in each of the clusters 

library(UpSetR)
library(tibble)

# Prepare the list of genes for each cluster
markers <- cluster_markers %>%
  group_by(cluster) %>%
  summarise(genes = list(gene)) %>%
  deframe()

# Create the UpSet plot
upset(fromList(markers), nsets = length(markers))

```

## Fast Gene Set Enrichement Analysis (with MultVI leiden column)
```{r}

mouse_brain <- read.csv(here::here("Data", "biorxiv_mouse_brain.csv"))
mbrain_genesets <- sapply(strsplit(mouse_brain$Optimally.Sized.Gene.List, "\\||="), function(i){
  i[-grep("ENS", i)]
})
names(mbrain_genesets) <- mouse_brain$Cell.Type.Name
mbrain_genesets <- mbrain_genesets[sapply(mbrain_genesets, length) != 0]

fc <- cluster_markers_filtered$avg_log2FC
names(fc) <- rownames(cluster_markers_filtered)
fc_cl <- split(fc, cluster_markers_filtered$cluster)


cluster_annotations <- lapply(fc_cl, function(i){
  res <- fgsea(mbrain_genesets, i)
  res <- res[order(res$pval),]
  res$pathway[1]
})
```


# Labelling clusters with cell type 
```{r}
RNA <- RenameIdents(RNA, cluster_annotations)
```

## Save cluster_annotations to add back into adata object for MultiVI 
```{r}
write.csv(as.data.frame(cluster_annotations), "cluster_annotations.csv", row.names = FALSE)
```

## Add meta data to seurat object 
```{r}
demo <- data.frame(ids = c("r3", "r4", "r15", "r16"),
                       drug = c("Saline", "Fentanyl", "Fentanyl", "Saline"), 
                       sex = c("Male", "Male", "Female", "Female"))
demo_10x <- RNA@meta.data %>% tibble::rownames_to_column("ids") %>% mutate(ids = gsub("_.*", "", ids))
demo_10x <- demo_10x %>% left_join(demo, by = "ids")
new_metadata <- demo_10x[, c("drug", "sex")]

RNA <- AddMetaData(object = RNA, metadata = new_metadata)

# Create a column in metadata called SampleID with same info as orig.ident 
RNA@meta.data$SampleID <- RNA@meta.data$orig.ident

# extract cell types 
cell_types <- as.character(RNA@active.ident)
RNA <- AddMetaData(object = RNA, metadata = cell_types, col.name = "cell_type")

# Join RNA count layers 
RNA[["RNA"]] <- JoinLayers(RNA[["RNA"]])

# Extract cell barcodes and genes from the SCT assay
cell_barcodes <- colnames(RNA)
gene_names <- rownames(RNA)

# Assign cell barcodes and gene names to the dimnames of the counts matrix
dimnames(RNA@assays$RNA@layers$counts) <- list(gene_names, cell_barcodes)
```

## Save GEX+ATAC seurat object 
```{r}
saveRDS(RNA, file = here::here("Data", 'GEX_ATAC_Seur.rds'))
```

