---
title: "Diff_Exp.Rmd"
output: html_document
date: "2024-05-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Differential Expression Analysis using Limma Voom 

## Load libraries 
```{r}
library(limma)
library(dplyr)
library(tidyr)
library(ggplot2)
library(EnhancedVolcano)
library(Seurat)
```

## Load data 
```{r}
rna <- readRDS("/Users/rishikadaswani/Desktop/rds/GEX_ATAC_Seur.rds")
```

## Add metadata 
```{r}
demo <- data.frame(ids = c("r3", "r4", "r15", "r16"),
                       drug = c("Saline", "Fentanyl", "Fentanyl", "Saline"), 
                       sex = c("Male", "Male", "Female", "Female"))
demo_10x <- rna@meta.data %>% tibble::rownames_to_column("ids") %>% mutate(ids = gsub("_.*", "", ids))
demo_10x <- demo_10x %>% left_join(demo, by = "ids")
```

## Extract counts 
```{r}
eset <- rna@assays$SCT@counts
ids <- unlist(sapply(strsplit(colnames(eset), "_"), function(i) i[[1]]))
sex <- factor(demo_10x[, "sex"], c("Male", "Female"))
drug <- factor(demo_10x[, "drug"], c("Saline", "Fentanyl"))
cell_labels <- as.character(Idents(rna))
```

## Limma Analysis 
### Limma has a built-in approach for analyzing repeated measures data using duplicateCorrelation(). The model can handle a single random effect, and forces the magnitude of the random effect to be the same across all genes.


```{r}
top <- lapply(names(table(cell_labels)), function(cell){
  keep <- cell_labels %in% cell
  counts <- eset[, keep]
  dat <- data.frame(sex = sex[keep],
                    drug = drug[keep],
                    subj = gsub("_.*", "", colnames(counts)))
  
  design <- model.matrix(~sex*drug, data = dat)
  vobj_tmp <- voom(counts, design, plot = FALSE)
  dupcor <- duplicateCorrelation(vobj_tmp, design, block = dat$subj)
  vobj <- voom(counts, design, plot = FALSE, block = dat$subj, correlation = dupcor$consensus)
  dupcor <- duplicateCorrelation(vobj, design, block = dat$subj)
  fitDupCor <- lmFit(vobj, design, block = dat$subj, correlation = dupcor$consensus)
  fitDupCor <- eBayes(fitDupCor)
  top_FvsM <- topTable(fitDupCor, coef = "sexFemale", adjust.method = "BH", number = nrow(fitDupCor))
  top_FvsM$comparison <- "Female_vs_Male"
  top_FvsM$gene <- rownames(top_FvsM)
  top_SvsF <- topTable(fitDupCor, coef = "drugFentanyl", adjust.method = "BH", number = nrow(fitDupCor))
  top_SvsF$comparison <- "Saline_vs_Fentanyl"
  top_SvsF$gene <- rownames(top_SvsF)
  top_int <- topTable(fitDupCor, coef = "sexFemale:drugFentanyl", adjust.method = "BH", number = nrow(fitDupCor))
  top_int$comparison <- "drug x sex interaction"
  top_int$gene <- rownames(top_int)
  top <- rbind(top_FvsM, top_SvsF, top_int)
  top$cell <- cell
top }) 
```

## Limma analysis (DupCor) with one cell type 
```{r}
cell = names(table(cell_labels))[1]
keep <- cell_labels %in% cell
  counts <- eset[, keep]
  dat <- data.frame(sex = sex[keep],
                    drug = drug[keep],
                    subj = gsub("_.*", "", colnames(counts)))
  
  design <- model.matrix(~sex*drug, data = dat)
  vobj_tmp <- voom(counts, design, plot = FALSE)
  dupcor <- duplicateCorrelation(vobj_tmp, design, block = dat$subj)
  vobj <- voom(counts, design, plot = FALSE, block = dat$subj, correlation = dupcor$consensus)
  dupcor <- duplicateCorrelation(vobj, design, block = dat$subj)
  fitDupCor <- lmFit(vobj, design, block = dat$subj, correlation = dupcor$consensus)
  fitDupCor <- eBayes(fitDupCor)
  top_FvsM <- topTable(fitDupCor, coef = "sexFemale", adjust.method = "BH", number = nrow(fitDupCor))
  top_FvsM$comparison <- "Female_vs_Male"
  top_FvsM$gene <- rownames(top_FvsM)
  top_SvsF <- topTable(fitDupCor, coef = "drugFentanyl", adjust.method = "BH", number = nrow(fitDupCor))
  top_SvsF$comparison <- "Saline_vs_Fentanyl"
  top_SvsF$gene <- rownames(top_SvsF)
  top_int <- topTable(fitDupCor, coef = "sexFemale:drugFentanyl", adjust.method = "BH", number = nrow(fitDupCor))
  top_int$comparison <- "drug x sex interaction"
  top_int$gene <- rownames(top_int)
  top <- rbind(top_FvsM, top_SvsF, top_int)
  top$cell <- cell
```

## Plot DEGs for the above cell type 
```{r}
top %>% filter(adj.P.Val < 0.001) %>% 
  group_by(cell, comparison) %>% 
  summarise(n = n()) %>% 
  ungroup %>% 
  complete(cell, comparison) %>% 
  mutate(comparison = factor(comparison, levels = c("drug x sex interaction", "Female_vs_Male", "Saline_vs_Fentanyl"))) %>% 
  ggplot(aes(x = reorder(cell, -n), y = n, fill = comparison)) + 
  geom_bar(stat = "identity", position = "dodge") + scale_y_log10() + 
  ylab("Number of differentially expressed genes (FDR<1%)") +   
  xlab("Cell-type") +   
  theme_classic() +   
  theme(legend.position = c(0.9,0.9)) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

top_filtered <- top %>% filter(adj.P.Val < 0.001)
top_filtered_dxs <- top_filtered[top_filtered$comparison == "drug x sex interaction", ]
```





