---
title: "Cosmx_Seurat.Rmd"
output: html_document
date: "2024-06-06"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Spatial Transcriptomics

## Load libraries 
```{r}
library(dplyr)
library(Seurat)
library(SeuratObject)
library(hdf5r)
library(ggplot2)
library(tidyr)
library(fgsea)
library(SeuratData)
library(SeuratDisk)
library(reticulate)
library(anndata)

future::plan("multisession", workers = 4)
options(future.globals.maxSize = 50000 * 1024^2) # for 50 Gb RAM

set.seed(1234)
```

## Load data 
```{r}
cosmx_eset <- read.csv(here::here("Data", "Run5618_RtBrain_tx_file.csv"))
cosmx_demo <- read.csv(here::here("Data", "fov_annotations.csv"))
```

## create counts 
```{r}
cosmx_counts0 <- cosmx_eset %>% 
  # filter(cell_ID != 0) %>% 
  # filter(CellComp == "Nuclear") %>% 
  mutate(cellid_fov = paste(cell_ID, fov, sep="_")) %>% 
  dplyr::select(cellid_fov, target) %>% 
  group_by(cellid_fov, target) %>% 
  summarise(n = n()) %>% 
  spread(cellid_fov, n, fill = 0)
cosmx_counts <- as.matrix(cosmx_counts0[, -1])
rownames(cosmx_counts) <- cosmx_counts0$target
```

## create seurat object using counts 
```{r}
cosmx <- CreateSeuratObject(counts = cosmx_counts, assay = "RNA",
                                 min.cells = 10, min.features = 10)
selectedfeatures <- rownames(cosmx)[-grep("Custom|NegPrb", rownames(cosmx))]
cosmx <- subset(
  x = cosmx,
  features = selectedfeatures
)
```

## Variance stablization and transformation 
```{r}
cosmx <- SCTransform(cosmx, assay = "RNA")
```

## Perform linear dimensional reduction 
```{r}
# perform PCA on scaled data 
cosmx <- RunPCA(cosmx, npcs = 50, features = rownames(cosmx))
```

## Run non-linear dimensionality reduction (UMAP)
```{r}
cosmx <- RunUMAP(cosmx, dims = 1:50)
```

## Cluster the cells 
```{r}
cosmx <- FindNeighbors(cosmx, reduction = "pca", dims = 1:50)
cosmx <- FindClusters(cosmx, resolution = 0.6)
```

## DimPlot 
```{r}
DimPlot(cosmx, reduction = "umap")
```

# Annotate using MapMyCells - Allen Brain Institute

## Save corrected counts to h5ad 
```{r}
use_condaenv(condaenv = '/Users/rishikadaswani/anaconda3/envs/cellanno', required = TRUE)

minimal_adata <- t(cosmx@assays$SCT$counts)

ad <- AnnData(
   X = minimal_adata,
   obs = data.frame(group = rownames(minimal_adata), row.names = rownames(minimal_adata)),
   var = data.frame(type = colnames(minimal_adata), row.names = colnames(minimal_adata))
 )

 write_h5ad(ad,here::here("Data", "cosmx_fent_counts.h5ad"), compression='gzip')
```

## The h5ad file gets inputted into the https://knowledge.brain-map.org/mapmycells/process

```{r}
#load output from above 
cellann0 <- read.csv(here::here("Data","cosmx_fent_counts_10xWholeMouseBrain(CCN20230722)_HierarchicalMapping_UTC_1718838739181", "cosmx_fent_counts_10xWholeMouseBrain(CCN20230722)_HierarchicalMapping_UTC_1718838739181.csv"), header = FALSE)
cellann <- cellann0[-c(1:5), ]
colnames(cellann) <- as.character(cellann0[5,])
```

```{r}

subclass <- na.omit(readxl::read_xlsx(here::here("Data", "cl.df_CCN202307220.xlsx"), sheet = "subclass_annotation"))
row.names(subclass) <- subclass$cell_set_accession.subclass
cc <- subclass[cellann$subclass_label, "subclass_label"]
rownames(cc) <- cellann$cell_id
cosmx@meta.data$aba_classification = cc[colnames(cosmx), ]

```

```{r}
aba_cellann <- table(cosmx$aba_classification)

celltypes_more20 <- aba_cellann[aba_cellann > 20] |> names()

keep_cellann <- cosmx@meta.data %>%
                filter(aba_classification$subclass_label %in% celltypes_more20)


cosmx <- subset(cosmx, cells = rownames(keep_cellann))
overlap <- table(cosmx$aba_classification, cosmx$seurat_clusters)
Heatmap(overlap/rowSums(overlap),
        cell_fun = function(j, i, x, y, w, h, col) { # add text to each grid
        grid.text(overlap[i, j], x, y)
    })
  

```

# Find Markers 
```{r}
Idents(cosmx) <- "aba_classification"
markers <- FindAllMarkers(cosmx,
                          assay = "SCT",
                          logfc.threshold = .1, 
                          only.pos = TRUE,
                          min.pct=0.1) %>% 
  group_by(cluster) %>% 
  dplyr::slice(1:2)
cluster_markers <- split(markers$gene, markers$cluster)
lapply(cluster_markers, length)

# Heatmap showing gene markers for each cell type 
DoHeatmap(cosmx, features = markers$gene, angle = 90, size = 2, vjust = 0.6, hjust = 0.2) + NoLegend() + theme(axis.text.y = element_text(size = 5.5, face = "bold")) 


# UMAP showing clusters with labelled cell types 
DimPlot(cosmx, reduction = "umap", label = TRUE, label.size = 4, repel = TRUE) + ggtitle("Spatial transcriptomics labelled clusters") + theme(plot.title = element_text(hjust = 0.5), legend.text = element_text(size = 14, face = "bold"))
```

## Add image data 
```{r}
data.dir <- here::here("07_Spatial_Analysis", "Rat_Brain")
fov = "rat_brain"
assay = "SCT"
data <- ReadNanostring(data.dir = data.dir, type = c("centroids", 
        "segmentations"))
segs <- CreateSegmentation(data$segmentations)
cents <- CreateCentroids(data$centroids)
segmentations.data <- list(centroids = cents, segmentation = segs)
coords <- CreateFOV(coords = segmentations.data, type = c("segmentation", 
    "centroids"), molecules = data$pixels, assay = assay)
cells <- intersect(Cells(x = coords, boundary = "segmentation"), 
    Cells(x = coords, boundary = "centroids"))
cells <- intersect(Cells(cosmx), cells)
coords <- subset(x = coords, cells = cells)
cosmx[[fov]] <- coords
cosmx
```


```{r}
ImageDimPlot(cosmx, fov = "rat_brain", axes = TRUE, cols = "glasbey",
             group.by = 'aba_classification', crop = TRUE) & NoLegend()
```

```{r}
tr <- cosmx
basal.crop <- Crop(tr[["rat_brain"]], x = c(-25000, -10000), y = c(326000, 333000))
tr[["rat_brain"]] <- basal.crop
DefaultBoundary(tr[["rat_brain"]]) <- "segmentation"
ImageDimPlot(cosmx, fov = "rat_brain", axes=TRUE, group.by = 'aba_classification', border.color = "black") + ggtitle("Cell types for each sample on spatial slide")
```

## load multiome seurat object 
```{r}
gex_atac <- readRDS(here::here("Data", "GEX_ATAC_Seur.rds"))
```

## Common cell types between GEX and spatial 
```{r}
common <- intersect(unique(gex_atac@meta.data$aba_classification),
          unique(cosmx@meta.data$aba_classification$subclass_label))
```

## Subset data for common cell types 
```{r}

subclass_level <- as.character(cosmx@meta.data$aba_classification$subclass_label)

cosmx$aba_classification <- subclass_level

cosmx_subset <- subset(cosmx, cells = Cells(cosmx)[cosmx@meta.data$aba_classification %in% common])

gex_atac_subset <- subset(gex_atac, cells = Cells(gex_atac)[gex_atac$aba_classification %in% common])
```

## Visualize common cell types on spatial slide 
```{r}
ImageDimPlot(cosmx_subset, fov = "rat_brain", axes = TRUE, cols = "glasbey",
             group.by = 'aba_classification', crop = TRUE) + ggtitle ("Common cell types between spatial and multiome on slide")

```


