---
title: "Top_table_analysis.Rmd"
output: html_document
date: "2024-06-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Top Table Analysis (heatmaps)

## Load libraries 
```{r}
library(ComplexHeatmap)
library(dplyr)
library(circlize)
library(tidyr)
```

## Load 3 top tables (gex, atac and gene activity)
```{r}
top_gex <- readRDS(here::here("Data", "topfil_GEX.rds"))
top_ga <- readRDS(here::here("Data", "topfil_GA.rds"))
```

## Getting counts for each dataset 
```{r}
#GEX counts 
top_gex <- top_gex %>% filter(p_adj.loc < 0.25)

# Process GEX dataset
df_gex <- top_gex %>%
  mutate(dataset = "gex") %>%
  select(cluster_id, dataset, logFC) %>%
  mutate(regulation = ifelse(logFC > 0, "UP", "DOWN"))

gene_counts <- df_gex %>%
  group_by(cluster_id, regulation) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = regulation, values_from = count, values_fill = 0) %>%
  mutate(dataset = "gex")

# ATAC Counts (filter for only one cell type as top_ga and top_gex only have DEGs for this one cell type)
#top_atac <- top_atac %>% filter(p_adj.loc < 0.20) %>% filter(cluster_id == "CB Granule Glut")

# Process ATAC dataset
#df_atac <- top_atac %>%
  #mutate(dataset = "atac") %>%
  #select(cluster_id, dataset, logFC) %>%
  #mutate(regulation = ifelse(logFC > 0, "UP", "DOWN"))

#atac_counts <- df_atac %>%
  #group_by(cluster_id, regulation) %>%
  #summarise(count = n()) %>%
  #pivot_wider(names_from = regulation, values_from = count, values_fill = 0) %>%
  #mutate(dataset = "atac")
  
# GA Counts 
top_ga <- top_ga %>% filter(p_adj.loc < 0.60)

# Process GA dataset
df_ga <- top_ga %>%
  mutate(dataset = "ga") %>%
  select(cluster_id, dataset, logFC) %>%
  mutate(regulation = ifelse(logFC > 0, "UP", "DOWN"))

ga_counts <- df_ga %>%
  group_by(cluster_id, regulation) %>%
  summarise(count = n()) %>%
  pivot_wider(names_from = regulation, values_from = count, values_fill = 0) %>%
  mutate(dataset = "ga")

```

## Plotting heatmap showing number of genes that are up- and down-regulated for each cell type within the three datasets
```{r}

# Combine datasets and ensure correct aggregation
combined_counts <- bind_rows(gene_counts, ga_counts)

# Aggregate counts to avoid duplicates
aggregated_counts <- combined_counts %>%
  pivot_longer(cols = c(UP, DOWN), names_to = "FC", values_to = "count") %>%
  mutate(count = ifelse(FC == "DOWN", -count, count)) %>%  # Convert DOWN counts to negative
  group_by(cluster_id, dataset, FC) %>%
  summarise(count = sum(count, na.rm = TRUE)) %>%  # Aggregate counts
  pivot_wider(names_from = dataset, values_from = count, values_fill = 0)

# Convert to matrix
heatmap_matrix <- as.matrix(aggregated_counts[, -c(1, 2)])  # Exclude cluster_id and FC for the matrix
rownames(heatmap_matrix) <- paste(aggregated_counts$cluster_id, aggregated_counts$FC, sep = "_")

# Define color palette for counts
col_fun <- colorRamp2(c(-max(abs(heatmap_matrix), na.rm = TRUE), 0, max(abs(heatmap_matrix), na.rm = TRUE)), c("lightblue", "white", "salmon"))

# Define unique colors for each cluster_id
cluster_ids <- unique(aggregated_counts$cluster_id)
cluster_colors <- setNames(rainbow(length(cluster_ids)), cluster_ids)

# Define row annotation colors for cluster_id and FC
row_ha = rowAnnotation(
  Cluster = factor(aggregated_counts$cluster_id),
  FC = factor(aggregated_counts$FC),
  col = list(Cluster = cluster_colors, FC = c("UP" = "salmon", "DOWN" = "lightblue"))
)

# Create heatmap for counts
Heatmap(
  heatmap_matrix,
  name = "Counts",
  col = col_fun,
  right_annotation = row_ha,
  cluster_rows = FALSE,
  show_row_names = FALSE,  # Remove row names
  show_column_names = TRUE,
  row_names_side = "left",
  column_names_side = "bottom",
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.1f", heatmap_matrix[i, j]), x, y, gp = gpar(fontsize = 10))  # Ensure numeric values
  },
  column_title = "Number of up and down regulated genes in each dataset"
)

```

## Heatmap (intersection of genes)
```{r}
# Define the categories
categories <- c("Only GA", "Intersect", "Only GEX")

# Create a combined list of all cell types
all_cell_types <- union(unique(top_gex$cluster_id), unique(top_ga$cluster_id))

# Create a dataframe with all cell types and initialize counts to zero
counts_df <- data.frame(
  cluster_id = all_cell_types,
  only_gex_genes = 0,
  intersect_genes = 0,
  only_ga_genes = 0
)

# Calculate the counts for each cell type present in top_gex
counts_gex <- top_gex %>%
  group_by(cluster_id) %>%
  summarise(
    only_gex_genes = sum(!gene %in% top_ga$gene),  # Comparing with top_ga
    intersect_genes = sum(gene %in% top_ga$gene)   # Intersecting with top_ga
  ) %>%
  ungroup()

# Calculate the counts for each cell type present in top_ga
counts_ga <- top_ga %>%
  group_by(cluster_id) %>%
  summarise(
    only_ga_genes = sum(!gene %in% top_gex$gene),   # Comparing with top_gex
    intersect_genes = sum(gene %in% top_gex$gene)   # Intersecting with top_gex
  ) %>%
  ungroup()

# Merge the counts into the dataframe with all cell types
counts_df <- left_join(counts_df, counts_gex, by = "cluster_id", suffix = c("", "_gex"))
counts_df <- left_join(counts_df, counts_ga, by = "cluster_id", suffix = c("", "_ga"))

# Replace NA values with 0
counts_df[is.na(counts_df)] <- 0

# Adjust the column names
counts_df <- counts_df %>%
  mutate(
    intersect_genes = intersect_genes_gex + intersect_genes_ga,  # Combine intersect counts
    only_gex_genes = only_gex_genes + only_gex_genes_gex,        # Combine only_gex counts
    only_ga_genes = only_ga_genes + only_ga_genes_ga             # Combine only_ga counts
  ) %>%
  select(cluster_id, only_gex_genes, intersect_genes, only_ga_genes)

# Convert to matrix
counts_matrix <- as.matrix(counts_df[, -1])
rownames(counts_matrix) <- counts_df$cluster_id

# Print counts_gex and counts_ga to check if counts are being calculated correctly
print(counts_gex)
print(counts_ga)

# Create the heatmap
Heatmap(counts_matrix,
        name = "Number of Genes",
        row_title = "Cell Types",
        column_title = "Gene Categories",
        show_row_names = TRUE,
        show_column_names = TRUE,
        column_names_side = "bottom")


```


## Find the intersecting genes 
```{r}
# Find unique intersecting genes for each cell type and store in a nested column
intersect_genes_df <- top_ga %>%
  filter(gene %in% top_gex$gene) %>%
  group_by(cluster_id) %>%
  summarise(intersect_genes = list(gene)) %>%
  ungroup()

# Check the resulting dataframe
print(intersect_genes_df)

```

# Bar plot 
```{r}
# Define the intersect genes for annotation
intersect_genes <- top_gex %>%
  filter(gene %in% top_ga$gene) %>%
  pull(gene)

# Prepare data for bar plot
bar_data <- data.frame(
  category = rep(c("Only GA", "Intersect", "Only GEX"), each = 1),
  count = c(sum(!top_ga$gene %in% top_gex$gene), 
            length(intersect_genes), 
            sum(!top_gex$gene %in% top_ga$gene)),
  genes = c(NA, paste(intersect_genes, collapse = ", "), NA)  # Add gene names for intersect
)

# Create the bar plot
ggplot(bar_data, aes(x = category, y = count, fill = category)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = genes), 
            vjust = -0.5, 
            size = 3.5, 
            hjust = 0.5, 
            position = position_stack(vjust = 0.5)) +
  labs(title = "Number of genes in each dataset",
       x = "Gene Categories",
       y = "Number of Genes") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```


# plot the intersecting gene
```{r}
# Define the intersecting genes
gene <- "Ugt8"

cells <- gex$aba_classification[gex$aba_classification == "CB Granule Glut"]


plot_data <- data.frame(gene = gex@assays$SCT@data[gene, cells], ga = ga@assays$SCT@data[gene, cells], drug = gex@meta.data[ cells, 'condition'])


p <- ggplot(plot_data, aes(x = ga, y = gene, color = drug)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +  # Add line of best fit
  labs(title = paste("Gene:", gene), x = "Gene Activity", y = "Gene Expression") +
  theme_minimal()

```

## Plot all of the intersecting genes 
```{r}
gex_df <- gex@assays$SCT@data[rownames(gex@assays$SCT@data) %in% intersect_genes, names(cells)] %>% 
  as.matrix()
ga_df <- ga@assays$SCT@data[rownames(ga@assays$SCT@data) %in% intersect_genes, names(cells)] %>% 
  as.matrix()

gex_df_final <- gex_df %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "genes") %>% 
  as_tibble() %>%
  pivot_longer(cols = !genes, names_to = "cells", values_to = "gex")

ga_df_final <- ga_df %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "genes") %>% 
  as_tibble() %>%
  pivot_longer(cols = !genes, names_to = "cells", values_to = "ga")

joined_df <- left_join(gex_df_final, ga_df_final, by = c("genes", "cells"))
metadata_df <- gex@meta.data[names(cells), ] %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "cells") %>% 
  as_tibble() %>% select(cells, condition)

final_df <- left_join(joined_df, metadata_df, by = "cells")

top6 <- final_df$genes %>% unique %>% tail(6)

final_df %>% 
  filter(genes %in% top6) %>%
  ggplot(aes(x = ga, y = gex, colour = condition)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +  # Add line of best fit
  facet_wrap(~genes) + 
  #labs(title = paste("Gene:", genes), x = "Gene Activity", y = "Gene Expression") +
  theme_minimal()

```

